---
title: "Detecting Anomalies in Financial Statements #2 - Accounting Anomalies, Benford distribution"
author: "Ramesh Calamur"
date: "January 17, 2016"
output: html_document
---
```{r "Environment setup", echo=FALSE,message=FALSE,warning=FALSE}

library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
#### Load financial facts
cash14 <- read.csv("cashflow_statement.csv")
balsh14 <- read.csv("balance_sheet.csv")
income14 <- read.csv("income_statement.csv")
ratios14 <- read.csv("ratios.csv")

### Load filings information
filings14 <- read.csv("filings.csv")
```

##2 - Feature selection: GAAP Accounting Anomalies, Benford distribution

* Financial statements must conform to GAAP (Generally Accepted Accounting Principles) standards

* Example: On 'Balancesheet', Assets must equal "Liability + Equity" 


```{r "Accounting Anomaly", echo=FALSE,message=FALSE,warning=FALSE}
### Accounting anomalies - Balancesheet
### By definition, balancesheet should balance "Total Assets" ==  "Liabilities" + "Equity" 
bAccAnom <- balsh14 %>% mutate( accAnomaly = (assetsTotal != liabilitiesAndStockholdersEquity )) 

bAccAnom %>% ggplot( aes(x=log(abs(assetsTotal)),y=log(abs(liabilitiesAndStockholdersEquity))))+
  geom_point(aes(color= accAnomaly)) + 
  xlab("Total Assets")+
  ylab("Liabilities + Equity")+
  ggtitle("Accounting Anomaly - Balancesheet (scaled)")

cat("Number of balancesheet anomalies:",sum(bAccAnom$accAnomaly,na.rm = TRUE))

```

# Benford's Law - financial data anomaly detection 

* Benford’s Law or the law of first digits: real-life numerical data conform to theoretical frequency distribution of first digits 

  + Benford’s law states that the first significant digits (FSDs) will be distributed in a very specific pattern. Take a look at the relative proportion with which digits from 1 to 9 appear in the first significant position among values in the data. 
  
* Benford's law is used by domain experts in forensic audits, accounting fraud detection 

* Features of the distribution of financial statement numbers that diverge from theoretical Benford distribution as a potential anomaly indicator 

* We explore balancesheet numerical data conformance to Benford distribution 

```{r, "Benford law", echo=FALSE,message=FALSE,warning=FALSE}
### Benford' law - indicators for anomalies
### Calculate Mean Absolute Deviation for Benford distribution of Balancesheet 
### Benford data set applies to each company statements (row) and numeric columns

library(benford.analysis)
n <-  nrow(balsh14)

balsh14.benford.MAD <- numeric(length = n)
for (i in 1:n){
  balsh14.stmtdat <- as.numeric(balsh14[i,5:ncol(balsh14)])
  balsh14.stmtdat[is.na(balsh14.stmtdat)] <- 0.01
  balsh14.benford.MAD[i] <- MAD(benford(as.numeric(balsh14.stmtdat)))
}  
## summary stats of MAD statistic of Benford distro of balancesheet 

print("Summary stats of MAD statistic of Benford distro of balancesheet ")
summary(balsh14.benford.MAD)


benf.MAD.crit <- 0.018 ## Empirical MAD value upper threshold for non-conforming 
qplot(seq_along(balsh14.benford.MAD),balsh14.benford.MAD,color = balsh14.benford.MAD > 0.018,
      ylab = "Mean Absolute Deviation",
      xlab = "Index", 
      main = "Benford distribution of Balancesheet data  - MAD test") +
  geom_hline(yintercept = 0.018,color="blue")


print("Begin Chi Squared test of conformance to Benford distribution")

#### Begin Chi Squared Benford test
#### using Table from http://www.itl.nist.gov/div898/handbook/eda/section3/eda3674.htm
#### Chi Squared critical value: Degrees of freedom: 42, 
library(BenfordTests)
n <-  nrow(balsh14)
balsh14.chisq.crit.value <- 58.124 # degress of freedom: 42, significance level 0.05

print("Degress of freedom in balancesheet data: 42, significance level 0.05.")
print("Chi-squared critical value is: 58.124")

balsh14.benford.chisq <- numeric(length = n)
for (i in 1:n){
  balsh14.stmtdat <- as.numeric(balsh14[i,5:ncol(balsh14)])
  balsh14.stmtdat[is.na(balsh14.stmtdat)] <- 0.01
  balsh14.benford.chisq[i] <- chisq.benftest(balsh14.stmtdat)$statistic[[1]]
}  

### The obtained value is greater than the critical value, so we can indeed say that 
### the data do not fit Benford’s Law
# Reference: http://blog.bigml.com/2015/05/15/detecting-numeric-irregularities-with-benfords-law/

sum(balsh14.benford.chisq > balsh14.chisq.crit.value)
qplot(seq_along(balsh14.benford.chisq),balsh14.benford.chisq,
      color = balsh14.benford.chisq > balsh14.chisq.crit.value,
      ylab = "Chi Squared",
      xlab = "Index", 
      main = "Benford distribution - Chi Squared test") +
  geom_hline(yintercept = balsh14.chisq.crit.value,color="blue")

#### Benford chart
print("Plot the baseline Benford distribution for comparison")
benDist <- pbenf(digits = 1) ## baseline Benford distribution
datDist <- signifd.analysis(balsh14.stmtdat,graphical_analysis=FALSE)$summary["freq",]
benDatDist <- as.data.frame(cbind(benDist,dataFreq = datDist))
ggplot(aes(x=seq_along(benDist),y=benDist),data=benDatDist)+
  geom_line(color="black", size = 1)+
  geom_line(y=datDist,color="red")+
  scale_x_continuous(breaks=seq(1,10,1))+
  xlab("Digit")+
  ylab("Frequency")+
  ggtitle("Benford distribution - Balancesheet Anomaly sample")
#### End Chi Squared Benford test
### Benford test on all of Balancesheet records

print("Benford test on all of Balancesheet values.")
balshNumValues <- as.vector(as.matrix(balsh14[,5:ncol(balsh14)]))
balshNumValues <- balshNumValues %>% na.omit()
balsh.datDist <- signifd.analysis(balshNumValues,graphical_analysis=FALSE)$summary["freq",]
balsh.benDatDist <- as.data.frame(cbind(benDist,dataFreq = balsh.datDist))
ggplot(aes(x=seq_along(benDist),y=benDist),data=balsh.benDatDist)+
  geom_line(color="black", size = 1)+
  geom_line(y=balsh.datDist,color="red")+
  scale_x_continuous(breaks=seq(1,10,1))+
  xlab("Digit")+
  ylab("Frequency")+
  ggtitle("Benford distribution - All Balancesheet values")
```


## Summary: GAAP Accounting Anomaly, Benford distribution as features

* Features that Financial Statement Anomalies GAAP Accounting anomaly:
  + GAAP Accounting anomalies as Assets == Liability+Equity
  + Financial statement numeric facts's conformance to Benford distribution
  
* Though these by themselves may not be issues, but could be "screening" factors 

* Further analysis is needed on correlation between these and material issues in Financial statements

