---
title: "Detecting Anamolies - Data exploration with ratios"
author: "Ramesh Calamur"
date: "January 18, 2016"
output: html_document
---

---
```{r "Environment setup", echo=FALSE,message=FALSE,warning=FALSE}

library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
#### Load financial facts
cash14 <- read.csv("cashflow_statement.csv")
balsh14 <- read.csv("balance_sheet.csv")
income14 <- read.csv("income_statement.csv")
ratios14 <- read.csv("ratios.csv")

### Load filings information
filings14 <- read.csv("filings.csv")
```

```{r "Helper functions", echo=FALSE,message=FALSE,warning=FALSE}
#### Helper Functions - User Defined Functions
#
#### user defined function for selecting numeric values and filtering NA values
#### Arguments: 
####        factsDF:  data frame with multiple variables as columns 
####        maxNACnt: threshold count of NAs values for the variable, if below this, the variable
####                  would be included for calculating correlation matrix
####  Returns:
####        numDFforCor - data frame of numeric variables for correlation
####
numericNAvars <- function(factsDF, maxNACnt){
  ### identify the numeric variables and count of NAs of the variables
  numNAvarDF <- factsDF %>% select(which(sapply(., is.numeric))) %>%
    summarise_each(funs(sum(is.na(.)))) %>% 
    gather(bvar,NAcnt) %>%
    filter( NAcnt > 0 & NAcnt < maxNACnt)
  ## from the data frame, select the numeric variables that have NAs below threshold   
  numDFforCor <-  factsDF %>% select(one_of(as.character(numNAvarDF$bvar))) %>%
    na.omit()
  rm(numNAvarDF)
  return(numDFforCor)
}

## Altman Z Score
## Arguments:
##    currentAssetsTotal:
##    currentLiabilitiesTotal:
##    assetsTotal:
##    retainedEarnings:
##    incomeBeforeInterestAndTax:  equivalent for EBIT
##    stockholderEquityTotal: equivalent for MarketCapAtEndOfPeriod
##    liabilitiesTotal:
##    revenueTotal
##  Returns:
##    zScore : computed Altman Z-score
##
altmanZscore <- function(currentAssetsTotal, currentLiabilitiesTotal, assetsTotal,
                         retainedEarnings, incomeBeforeInterestAndTax,
                         stockholderEquityTotal,liabilitiesTotal,revenueTotal ){
    workingCapital <- currentAssetsTotal - currentLiabilitiesTotal
    zScore <- (1.2 * (workingCapital / assetsTotal) +
               1.4 * (retainedEarnings / assetsTotal) +
               3.3 * (incomeBeforeInterestAndTax / assetsTotal) +
               0.6 * (stockholderEquityTotal / liabilitiesTotal) +
               0.99 * (revenueTotal / assetsTotal) )
  return (zScore)
}
#### End of User Defined Functions
```

## Feature selection # 1 - Standard Financial Ratios 

* Select from the "Standard" financial ratios for anomaly detection
  
  + Correlation matrix of financial ratios to select ratios
  
  + Efficiency ratios: assetTurnover, fixedAssetTurnover, receivablesTurnover, inventoryTurnover were selected

```{r "Explore correlations", echo=FALSE,message=FALSE,warning=FALSE}
### Feature selection - find correlation
library(caret)
set.seed(2016)

### get a subset of numeric variables to find correlation matrix
## select only the numeric columns: select(which(sapply(., is.numeric))) 
### Count of NAs in the balance sheet data set

### Balancesheet facts - consider numeric variables that have less than 200 NAs
balShforCor <- numericNAvars(balsh14,200)
### Correlation matrix for Balancesheet data
balCor <- cor(balShforCor)
#knitr::kable(balCor) # print correlation matrix for balancesheet facts

#### Correlation matrix for income statement
### now extract the subset data for numeric variables selected for correlation
incomeforCor <- numericNAvars(income14,1000)
### Correlation matrix for Income statement data
incomeCor <- cor(incomeforCor)
#knitr::kable(incomeCor)
#### Correlation matrix for cash flow
### now extract the subset data for numeric variables selected for correlation
cashforCor <- numericNAvars(cash14,10000)
### Correlation matrix for cash statement data
cashCor <- cor(cashforCor)
#knitr::kable(cashCor)

## Correlation matrix for ratios 
### now extract the subset data for numeric variables selected for correlation
ratioforCor <- numericNAvars(ratios14,200)
### Correlation matrix for Balancesheet data
ratioCor <- cor(ratioforCor)
knitr::kable(ratioCor)
```

```{r "Ratios by SIC", echo=FALSE,message=FALSE,warning=FALSE}
### Get the assignedSic code to rations, by joining ratios with filings
ratios14NSic <- ratios14 %>% 
  left_join(.,filings14, by = c("accessionNumber" = "accessionNumber", "cikNumber" = "cikNumber")) %>%
  select(accessionNumber:dividendCoverage,assignedSic)
  
### explore the data now
#### binning SIC categories
ratios14NSic$sicCat <- cut(ratios14NSic$assignedSic, breaks = c(1,1000,1500,1800,2100,2400,3569,4000,5000,5200,6000,8000,9000,10000),
                           labels=c("Agriculture","Mining","Construction","Food","Textiles","Industrials","Manufacturing","Transportation","Wholesale","Retail","Finance","Services","Other") )

ggplot(aes(x=assignedSic,y=assetTurnover),data=subset(ratios14NSic,!is.na(assetTurnover)))+
  geom_jitter()+
  coord_cartesian(ylim = c(0,20))+
  facet_wrap(~sicCat)+
  ggtitle("Asset Turnover Ratio by SIC categories")


```

## Feature selection # 2 - Alternatives to Standard Financial Ratios
## Computed Altman Z-score - anomaly risk score based on multiple financial ratios

* For easier communication of anomaly indicator, a risk score factor of considering multiple ratios can be considered

* Altman Z-score, first published in 1968, reliably sued  as predictor for backruptcy  manufacturing companies

* Z-score = 0.012 x (Working Capital / Total Assets) +
                0.014 x (Retained Earnings / Total Assets) +
               0.033 x (Earnings before Interest and Taxes) / Total Assets) +
              0.006 x (Market Value of Equity / Book value of Total Liabilities)
               0.999 x (Sales / Total Assets)

* Average for bankrupt companies: -0.25, non-bankrupt companies: +4.48

```{r "Altman Z-score", echo=FALSE,message=FALSE,warning=FALSE}
#### join income statement and cashflow to calculate Altman Z-scores
incNbal <- income14 %>% 
        left_join(.,balsh14) %>% 
          left_join(.,filings14, by = c("accessionNumber" , "cikNumber" )) %>%
            select(1:5,currentAssetsTotal, currentLiabilitiesTotal, assetsTotal,
              retainedEarnings, incomeBeforeInterestAndTax,
              stockholderEquityTotal,liabilitiesTotal,revenueTotal,assignedSic) %>%
              na.omit() %>% 
            mutate( zScore = altmanZscore(currentAssetsTotal, currentLiabilitiesTotal, assetsTotal,
                              retainedEarnings,incomeBeforeInterestAndTax,
                              stockholderEquityTotal,liabilitiesTotal,revenueTotal) )
#####
## Summary statistics of Z-score
summary(incNbal$zScore)
###
#### binning SIC categories and ratio/score to plot by
binBySicPlot <- function(dfWithSic, zScore, pTitle){
  
  dfWithSic <- mutate(dfWithSic, zScoreAnom = zScore > 10) 
  dfWithSic$sicCat <- cut(dfWithSic$assignedSic, breaks = c(1,1000,1500,1800,2100,2400,3569,4000,5000,5200,6000,8000,9000,10000),
                             labels=c("Agriculture","Mining","Construction","Food","Textiles","Industrials","Manufacturing","Transportation",
                                      "Wholesale","Retail","Finance","Services","Other") )
 
 ggplot(aes(x=assignedSic,y=zScore,color=zScoreAnom),data=subset(dfWithSic))+
    geom_jitter()+
    coord_cartesian(ylim = c(0,20))+
    facet_wrap(~sicCat)+
    ggtitle(pTitle)
}

binBySicPlot(dfWithSic = incNbal,zScore = incNbal$zScore, pTitle = "Z-score by SIC")
#### Plot Z-Score by SICs
zScoreAnomAll <- incNbal$zScore > 10 
ggplot(aes(x=cikNumber,y=zScore,color = zScoreAnomAll),data=subset(incNbal))+
  geom_jitter()+
  coord_cartesian(ylim = c(0,100))+
  ggtitle("Z-score by CIK")

```

##Computed Altman z-Score as Financial Statement anomaly risk factor

* Altman zScore can be used as anomaly risk factor

* zScore threshold for anomaly varies by industry sectors. In the above analysis, zScore threshold of 10 are exhibited in Mining, Industrials, Manufacturing, Wholesale, Retail and Finance sectors.

* Further analysis should refine zScore threshold for individual sectors or using clustering to group companies of similar financial profile, instead of using one across industry sectors
